Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-05 E-Hypergraphs-LICS24.tex
Section 5

CHRIS: REMARK THAT WE CANNOT DEAL WITH SIGNATURES WITH $f:0 \to 0$ MORPHISMS, AND WHY!
As extended cospans have a more general notion of interface we need to make some adjustments to DPOI rewriting.
The main issue is that internal interfaces may not be preserved during rewriting, for example, if the right-hand side of the rewrite rule has more hierarchy than the left-hand side: it affects both the eventual e-hypergraph and the possible pushout complement.

% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

@@

To account for such changes in the interfaces we put some additional constraints on DPOI rewriting for the case of $\catname{EHyp_{\Sigma}}$.
Consider a DPOI diagram with extra constraints below which encodes
the notion that e-hypergraph $G$ rewrites into $H$ with external interfaces $n_{ext}+m_{ext}$.
Note the new interfaces and how we use the same trick to go from cospan $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$ to $G \xleftarrow{} g_{i} + g_{o} \xleftarrow{} n_{ext} + m_{ext}$.
The original DPO square is depicted in blue.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

@@

% The parts of the original DPOI square is shown in blue.
% For example, the left-hand side $L$ of a rewrite rule
%  $\langle L, R \rangle$ which originally was some cospan
%  $i \xrightarrow{} L \xleftarrow{} j$ in $\MdaCospans$ 
%  now is a cospan
%  $x_{ext} \xrightarrow{} l_i \to L \xleftarrow{} l_o \xleftarrow{} y_{ext}$
% from $\WellTypedMdaEcospans$ where $l_i, l_o$ are whole input and output interfaces respectively.
% The top half of this diagram says that the outermost interfaces
%  should coincide for $L$ and $R$.
% The bottom half puts the same restrictions on $G$ and $H$.
% For the pushout to exist we impose our assumptions~\ref{pushout:assumptions} on the correspoding arrow.

We then also adjust the definition of a boundary complement.

@@

For monogamous extended cospans $x_{ext} \xrightarrow{} l_{i} L \xleftarrow{} l_{o} \xleftarrow{} y_{ext}$ and $n_{ext} \xrightarrow{} g_{i} \xrightarrow{} G \xleftarrow{} g_{o} \xleftarrow{} m_{ext}$ and mono $f : L \to G$, a pushout complement $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ as depicted below
\[
\scalebox{0.75}{

@@

\begin{definition}[Convex down-closed match] 
    We call $m : L \to G$ in $\catname{EHyp_{\Sigma}}$ a convex down-closed match if

    \begin{enumerate}
        \item $m$ is monomorphism
        \item $m(L)$ is a convex down-closed subgraph of $G$
    \end{enumerate}
\end{definition}
\

@@

if 
\begin{align*}    
    G \xleftarrow{\langle a_2,b_2 \rangle} g_i + g_{o}& \xleftarrow{\langle a_1, b_1 \rangle} n_{ext} + m_{ext} \\ &\Rrightarrow^{+}\\ H \xleftarrow{\langle a_2,b_2 \rangle} h_i + h_{o}& \xleftarrow{\langle a_1, b_1 \rangle} n_{ext} + m_{ext}
\end{align*}

under a rewrite rule
\[
    \langle l_{i} + l_{o} \xrightarrow{\langle a_2, b_2 \rangle} L \xleftarrow{\langle a_1, b_1 \rangle} x_{ext} + y_{ext},
    r_{i} + r_{o} \xrightarrow{\langle a_2, b_2 \rangle} R \xleftarrow{\langle a_1, b_1 \rangle} x_{ext} + y_{ext} 
    \rangle
\]

@@

So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{Frobenius} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined an $[-]$ is a homomorphism with respect to sources and targets.

Recall that $<_{c}$ is essentially a transitive clousure of $<_{c}^{\mu}$ and hence we can first define 
    \[<_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{E_{X} + E_{Y}}{\sim}}[u]_{E})

@@

        <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]_{V})

>>>>

CHRIS: REMARK THAT WE CANNOT DEAL WITH SIGNATURES WITH $f:0 \to 0$ MORPHISMS, AND WHY!
As extended cospans have a more general notion of interface we need to make some adjustments to DPOI rewriting.
The first thing is that we need to account for internal interfaces after all.
The main issue is then that internal interfaces may not be preserved during rewriting.
To illustrate this, consider the following rewrite rule
\[\langle x \xrightarrow{} x' \xrightarrow{} \mathcal{L} \xleftarrow{} y' \xleftarrow{} y, x \xrightarrow{} x'' \xrightarrow{} \mathcal{R} \xleftarrow{} y'' \xleftarrow{} y \rangle \]
such that $x' \not \cong x''$ and $y' \not \cong y''$ and consider the following e-hypergraph to perform the rewrite on
\[
    n \xrightarrow{} n' \xrightarrow{} \mathcal{G} \xleftarrow{} m' \xleftarrow{} m
\] such that there is a matching $m : \mathcal{L} \to \mathcal{G}$.
When removing the occurence of $\mathcal{L}$ from $\mathcal{G}$, e.g. by computing the pushout complement, the internal interface of $\mathcal{G}$ should be modified as the verticies of $\mathcal{L}$ corresponding to strict internal interface of $\mathcal{L}$ (and hence to strict internal interface of $\mathcal{G}$) should be removed from $\mathcal{G}$ and therefore from $\mathcal{G}$'s internal interface.
Then, when gluing $\mathcal{R}$ into the created context, we also introduce strict internal interfaces of $\mathcal{R}$ into this context.
% The input interface on the left-hand side of the rule is $\{1, 2\}$, while the input interface on the right-hand side of the rule is $\{1,2,5,6,7,8\}$.
% Note that the outermost interfaces are preserved.

@@

To account for such changes in the interfaces we put some additional structure on DPOI rewriting for the case of $\catname{EHyp({\Sigma})}$.
Consider a DPOI diagram with extra constraints below which encodes
the notion that e-hypergraph $G$ rewrites into $H$ with external interfaces $n_{ext}+m_{ext}$.
Note the addition of internal interfaces and how we use the same trick to go from cospan $n_{ext} \xrightarrow{} g_i \xrightarrow{} G \xleftarrow{} g_o \xleftarrow{} m_{ext}$ to $G \xleftarrow{} g_{i} + g_{o} \xleftarrow{} n_{ext} + m_{ext}$.
The original DPO square is depicted in blue.
% \[\begin{tikzcd}
% 	&&&& {i + j} \\
% 	{l_i+l_o} && L &&&& R && {r_i+r_o} \\
% 	\\
% 	{g_i+g_o} && {G^{\urcorner}} && C && {^{\ulcorner}H} && {h_i+h_o} \\
% 	\\
% 	&&&& {n+m}
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=2-7]
% 	\arrow["m", color={rgb,255:red,214;green,92;blue,92}, from=2-3, to=4-3]
% 	\arrow["{[i_c,j_c]}", color={rgb,255:red,214;green,92;blue,92}, from=1-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=2-7, to=4-7]
% 	\arrow["{[n_c,m_c]}"', color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-5]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-7]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=4-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-3]
% 	\arrow[color={rgb,255:red,214;green,92;blue,92}, from=6-5, to=4-7]
% 	\arrow[from=2-1, to=2-3]
% 	\arrow["{[f_i,f_o]}", from=2-1, to=4-1]
% 	\arrow[from=4-1, to=4-3]
% 	\arrow[from=2-9, to=2-7]
% 	\arrow[from=4-9, to=4-7]
% 	\arrow["{[h_i,h_o]}"{description}, from=2-9, to=4-9]
% 	\arrow[from=6-5, to=4-9]
% 	\arrow[from=6-5, to=4-1]
% 	\arrow["{[i_l,j_l]}"{description}, from=1-5, to=2-1]
% 	\arrow["{[i_r,j_r]}"{description}, from=1-5, to=2-9]
% \end{tikzcd}\]

@@

Then, we explicitly show how the internal interfaces of the pushout complement are constructed.

@@

For monogamous extended cospans $x_{ext} \xrightarrow{} l_{i} \xrightarrow{} L \xleftarrow{} l_{o} \xleftarrow{} y_{ext}$ and $n_{ext} \xrightarrow{} g_{i} \xrightarrow{} G \xleftarrow{} g_{o} \xleftarrow{} m_{ext}$ and mono $f : L \to G$, a pushout complement $x_{ext} + y_{ext} \to C \to G^{\urcorner}$ as depicted below
\[
\scalebox{0.75}{

@@

\begin{definition}[Convex down-closed match] 
    We call $m : L \to G$ in $\catname{EHyp(\Sigma)}$ a convex down-closed match if

    \begin{enumerate}
        \item $m$ is monomorphism
        \item $m(L)$ is a convex down-closed subgraph of $G$
    \end{enumerate}
\end{definition}

The whole process of rewriting including the construction of the internal interfaces of the resulting e-hypergraph is summarised in the following definition.

@@

if 
\begin{align*}    
    G \xleftarrow{[ a_2,b_2 ]} g_i + g_{o}& \xleftarrow{[a_1, b_1]} n_{ext} + m_{ext} \\ &\Rrightarrow^{+}\\ H \xleftarrow{[a_2,b_2]} h_i + h_{o}& \xleftarrow{[a_1, b_1]} n_{ext} + m_{ext}
\end{align*}

under a rewrite rule
\[
    \langle l_{i} + l_{o} \xrightarrow{[a_2, b_2]} L \xleftarrow{[a_1, b_1]} x_{ext} + y_{ext},
    r_{i} + r_{o} \xrightarrow{[a_2, b_2]} R \xleftarrow{[a_1, b_1]} x_{ext} + y_{ext} 
    \rangle
\]

@@

So far, the construction matches the construction of the pushout in $\catname{Hyp_{\Sigma}}$ which can be found in~\cite{Frobenius} and hence $s_{\sfrac{X+Y}{\sim}}$ and $t_{\sfrac{X+Y}{\sim}}$ are well-defined and $[-]$ is a homomorphism with respect to sources and targets.

Recall that $<$ is essentially a transitive clousure of $<^{\mu}$ and hence we can first define 
    \[<_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{E_{X} + E_{Y}}{\sim}}[u]_{E})

@@

        <_{\sfrac{X+Y}{\sim}}^{\mu}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]_{V})
